---
title: "Assignment 1"
author: "Beatrix Jones"
date: "7/23/2019"
output: html_document
---

The file "fungal-with-order.csv" contains measurements of small molecules (*metabolites*) from 118 different samples. 

The first four columns contain information about the sample, and the remaining ones contain the metabolite measurements. 

Most samples have fungus growing in them (18 different types), but some have broth only.  This is specified in the "isolate" column. Half the samples have also had a dose of short chain fatty acids  added, in order to understand the metabolic changes induced. This information is in the  "Condition" column. The ID column combines these two bits of information, and the order column specifies the order in which the measurements were taken.  

```{r, echo=FALSE}
read.csv("fungal-with-order.csv")->fungal
```

1. Make a boxplot that shows each of the metabolite columns (starting column 5), and compute the standard deviation of each of these columns (using the "apply" command) and make a histogram. Comment on whether the metabolites are on the same scale, and suggest what would happen if principal components was performed without scaling the data.

```{r, fig.width=6, fig.height=10}
boxplot(fungal[,5:337], ylab="Metabolite", xlab="Intensity", horizontal=TRUE, names=c(1:333))
axis(1)
```


```{r}

fungal.sd<-apply(fungal[,5:337],2,sd)
hist(fungal.sd, main="Standard deviations of the metabolites", xlab="Standard Deviation")
```
*The metabolites do not seem to be on the same scale: they have a huge range of standard deviations, with a few very large and many very small. If we do not scale the data, the principal components will be dominated by the high variance metabolites.*

2. Make a visualization of the correlation matrix of the metabolite measurements.  Are there groups of highly correlated metabolites?  Explain what features you are basing your conclusion on. 

```{r, fig.height=10, fig.width=10}
library(ggplot2)
library(reshape2)
names(fungal)[5:337]<- as.character(1:333)  
funcor<-cor(fungal[,5:337])
FcorMelt<-melt(funcor) # have a look---what has this done? 
names(FcorMelt)<-c("Metabolite1", "Metabolite2", "Correlation")
p<-ggplot(data=FcorMelt, aes(x=Metabolite1, y=Metabolite2, fill=Correlation)) + geom_tile() 
p+labs(x=NULL, y=NULL)+labs(title="Correlations Between Metabolites", x="Metabolite 1", y="Metabolite 2")
```

*High correlations are indicated by bright blue (as on the diagonal where the correlation is one).  We see many off diagonal elements in this shade, indicating pairs or blocks of metabolites with high correlations.  We also see a few very dark blocks, indicating strong negative correlations.*

3.  Perform principle components analysis of the scaled data and produce a scree plot.  Also give output showing what percentage of variability each of the first 10 principle components accounts for. How many priciple components would you use based on the "elbow" criteria?    

```{r}
fun.pr<-prcomp(fungal[, 5:337], scale=TRUE)
plot(fun.pr, main="Variance of each principal component")
round(fun.pr$sdev[1:10]^2/sum(fun.pr$sdev^2),2)
```
*I would say the "elbow" is at three, meaning we should use the first two principle components.  There is also an argument for either 3 or four principal components.*


4. Produce plots of the scores on the first two principal components.  First, use colors and/or symbols to indicate the fungus type.  Make a second plot using colors and/or symbols to indicate condition.  
```{r, fig.height=10, fig.width=10}
#Fine (and easier) if you just you symbols here rather than symbols+colors. 
 plot(fun.pr$x[as.numeric(fungal$Isolate) < 9, 1:2],
      col = fungal$Isolate[as.numeric(fungal$Isolate) < 9], 
      pch = 16, 
      xlim = range(fun.pr$x[,1]), ylim=range(fun.pr$x[,2]))


points(fun.pr$x[as.numeric(fungal$Isolate) < 17 & as.numeric(fungal$Isolate)>8 ,1:2],
       col=fungal$Isolate[as.numeric(fungal$Isolate) > 8  &
                            as.numeric(fungal$Isolate)< 17 ], 
       pch=8)

points(fun.pr$x[as.numeric(fungal$Isolate)> 16, 1:2],
       col=fungal$Isolate[as.numeric(fungal$Isolate)> 16] , pch=17)

legend("bottomright", 
       pch=c(rep(16,8), rep(8,8), rep(17,3)), 
       col=c(1:8, 1:8, 1:3), 
       legend=levels(fungal$Isolate))

```

```{r}
plot(fun.pr$x[,1:2], col=fungal$Condition)
legend("bottomright", col=c(1:2), pch=1, legend=levels(fungal$Condition))
```

5. Describe any clusters you see in the plots produced in (4).  Do they correspond to the fungus and/or condition?  Identify any outlying points.  Where do they fall in the order of measurements (ie could they be accounted for by "start up" problems, or problems at the end of the experiment?)

*We might (roughly) describe 3 clusters*

1. PC1 between -10 and 0, pc2 between 0 and 10; 

2. PC1 between 0 and 20

3. PC2 between 0 and -20

*In most cases all samples of a particular fungus (isolate) like in the same cluster.  All the clusters have a mix of the Control and d_SCFA conditions, anthough the third cluster (with negative PC2 values) seems to have a higher proprtion of the SCFA condition.*

*There are four outliers, two with very high PC1 values (blue stars in the first plot), and two with very negative PC (a black dot and pink star).  The blue stars are the first two samples run, so might be attributed to start up problems. The black dot is the last sample run; again there may be issues around the order.  The pink star is the 20th sample run, so there is not a clear issue to investigate around order.* 
```{r}
which(fun.pr$x[,1]> 20)->index1
which(fun.pr$x[,2]< -30)->index2
fungal$order[index1]
fungal$order[index2]
```